---
title: "<small><i>MSc Ecology and Data Science Thesis</i></small> <br> Functional traits of trees through forest successional time <br> - </br> <small> Data exploration / Explorative Analyses </br>"
author: "Merlin Weiss"
date: "Status: `r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    bibliography: dissexp.bib
---

```{r setup, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}

rm(list = ls()) # clean environment
options(scipen = 999) # disable scientific notation
set.seed(123) # set seed for reproducibility

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Volumes/ritd-ag-project-rd01pr-dmayn10/merlin")

library(DT)
library(cowplot)
library(knitr)
library(kableExtra)
library(htmltools)
library(purrr)
library(tidyverse)

rd_pathway <- "/Volumes/ritd-ag-project-rd01pr-dmayn10" # path to UCL research drive

## Read most recent data
fia_clean <- list.files("/Volumes/ritd-ag-project-rd01pr-dmayn10/merlin/data/fia_traits", full.names = TRUE) %>%
	file.info() %>%
	as_tibble(rownames = "file") %>%
	arrange(desc(mtime)) %>%
	slice(1) %>%
	pull(file) %>%
	read_csv() %>% 
	mutate(
		PID = as.character(PID),
		state = as.factor(state),
		standage = as.integer(standage),
		INVYR = as.integer(INVYR),
		FORTYPCD = as.integer(FORTYPCD),
		foresttype = as.factor(FORTYPCD),
		biom = as.factor(biome),
	  ownership = as.factor(ownership),
		managed = as.factor(managed),
		ll_id = as.character(ll_id),
	  across(starts_with("wmean_"), as.numeric),
		across(starts_with("total_"), as.numeric),
	  across(starts_with("fun_"), as.numeric),
		resource_use_score = as.numeric(resource_use_score),
		annual_mean_temperature = as.numeric(annual_mean_temperature),
		annual_precipitation = as.numeric(annual_precipitation),
		isothermality = as.numeric(isothermality),
		max_temperature_of_warmest_month = as.numeric(max_temperature_of_warmest_month),
		mean_diurnal_range = as.numeric(mean_diurnal_range),
		mean_temperature_of_coldest_quarter = as.numeric(mean_temperature_of_coldest_quarter),
		mean_temperature_of_driest_quarter = as.numeric(mean_temperature_of_driest_quarter),
		mean_temperature_of_warmest_quarter = as.numeric(mean_temperature_of_warmest_quarter),
		mean_temperature_of_wettest_quarter = as.numeric(mean_temperature_of_wettest_quarter),
		min_temperature_of_coldest_month = as.numeric(min_temperature_of_coldest_month),
		precipitation_seasonality = as.numeric(precipitation_seasonality),
		precipitation_of_coldest_quarter = as.numeric(precipitation_of_coldest_quarter),
		precipitation_of_driest_month = as.numeric(precipitation_of_driest_month),
		precipitation_of_driest_quarter = as.numeric(precipitation_of_driest_quarter),
		precipitation_of_warmest_quarter = as.numeric(precipitation_of_warmest_quarter),
		precipitation_of_wettest_month = as.numeric(precipitation_of_wettest_month),
		precipitation_of_wettest_quarter = as.numeric(precipitation_of_wettest_quarter),
		temperature_annual_range = as.numeric(temperature_annual_range),
		temperature_seasonality = as.numeric(temperature_seasonality)
		)
```

# Background 

Here, we explore the cleaned FIA data with the aim to understand how functional traits change through forest successinal time, and to make predictions for given climate change scenarios. The data is structured to the plot level. We have the following varibales available:

- **Standage**

- **Resource Use Score**, a metric which puts every plot on a spectrum from 0 to 1, where 0 translates to a highly acquisitive resource use strategy, and 1 to a highly conservative strategy:

			- resource_use_score

- **Functional diversity indices**:

		- fun_disp (functional dispersion)
		- dun_div  (functional diversity)
		- fun_even  (functional evenness)

- **Trait meaans** (weighted by basal area scaled to trees per acre):

		- wmean_wood_density
		- wmean_bark_thickness
		- wmean_conduit_diam
		- wmean_leaf_n
		- wmean_specific_leaf_area
		- wmean_seed_dry_mass
		- wmean_shade_tolerance
		- wmean_dia
		- wmean_height

- Plot sums of **forest stand metrics**:

		- total_carbon_ag (total carbon above the ground)
		- total_biomass
		- total_basal_area

- Categorical **stand variables**:

		- biome
		- foresttype
			
- **Climate** variables: 

		- annual_mean_temperature
		- annual_precipitation
		- isothermality
		- max_temperature_of_warmest_month
		- mean_diurnal_range
		- mean_temperature_of_coldest_quarter
		- mean_temperature_of_driest_quarter
		- mean_temperature_of_warmest_quarter
		- mean_temperature_of_wettest_quarter
		- min_temperature_of_coldest_month
		- precipitation_seasonality
		- precipitation_of_coldest_quarter
		- precipitation_of_driest_month
		- precipitation_of_driest_quarter
		- precipitation_of_warmest_quarter
		- precipitation_of_wettest_month
		- precipitation_of_wettest_quarter
		- temperature_annual_range
		- temperature_seasonality

See the following table for min, max, and mean of numeric variables: 

```{r numeric variables, echo=FALSE, warning=FALSE, message=FALSE}
generate_numerical_summary <- function(dat) {
  num_vars <- dat %>% select(where(is.numeric))
  
  # Compute summary statistics
  summary_df <- num_vars %>%
    summarise(across(everything(), list(
      min = ~min(.),
      max = ~max(.),
      mean = ~mean(.)
    ))) %>%
    pivot_longer(everything(), names_to = c("variable", ".value"), names_pattern = "(.*)_(.*)") %>%
    mutate(min = round(min, digits = 3),
           max = round(max, digits = 3), 
           mean = round(mean, digits = 3))
           
   DT::datatable(summary_df, escape = FALSE, options = list(lengthMenu = c(5, 20, 50), pageLength = 5))
}

generate_numerical_summary(fia_clean)
```

# Climate variables 

Check for correlations

```{r cor matrix climate vars, echo=TRUE, fig.width=12, fig.height=10, fig.cap = "Correlation matrix of climate variables"}

require(corrplot)
require(caret)

clim <- fia_clean %>%
	# Select climate variables
	select(annual_mean_temperature:temperature_seasonality) %>%
	# Calculate correlation matrix
	cor()

# Check correlation plot
corrplot(clim, method = "circle", type = "upper", tl.col = "black", tl.srt = 45)

# Find highly correlated variables to be dropped
drop <- findCorrelation(clim, cutoff = 0.7, verbose = TRUE)
drop <- colnames(clim)[drop]

fia_clean <- fia_clean %>%
	dplyr::select(-all_of(drop))

```



# Resource Use Score

```{r}
# Check distribution 
ggplot(fia_clean, aes(x = resource_use_score)) +
	geom_histogram(binwidth = 0.05, fill = 'blue', color = 'black', alpha = .8) +
	theme_bw() +
	labs(title = 'Distribution of Resource Use Score', x = 'Resource Use Score', y = 'Frequency')

fia_clean %>%
	filter(complete.cases(biome) & biome != "Mangroves") %>%
	ggplot(aes(x = resource_use_score, group = biome)) +
	geom_histogram(binwidth = 0.05, fill = 'blue', color = 'black', alpha = .8) +
	facet_wrap(~biome, scale = "free") +
	theme_bw() +
	labs(title = 'Distribution of Resource Use Score per Biome', x = 'Resource Use Score', y = 'Frequency')

fia_clean %>%
		mutate(
		foresttype = case_when(
			    FORTYPCD >= 101 & FORTYPCD <= 199 ~ "Coniferous",
			    FORTYPCD >= 201 & FORTYPCD <= 299 ~ "Deciduous",
			    FORTYPCD >= 301 & FORTYPCD <= 399 ~ "Mixed",
			    TRUE ~ "Non-stocked/Other"
  )) %>%
	ggplot(aes(x = resource_use_score, group = foresttype)) +
	geom_histogram(binwidth = 0.05, fill = 'blue', color = 'black', alpha = .8) +
	facet_wrap(~foresttype, scale = "free") +
	theme_bw() +
	labs(title = 'Distribution of Resource Use Score per Forest Type', x = 'Resource Use Score', y = 'Frequency')

```



