---
title: "<small><i>MSc Ecology and Data Science Thesis</i></small> <br> Functional traits of trees through forest successional time <br> - </br> <small> Paper narrative, data exploration, preparation, and analysis </br>"
author: "Merlin Weiss"
date: "Status: `r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    bibliography: dissexp.bib
---

```{r setup, include=FALSE}

rm(list = ls()) # clean environment 

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Volumes/ritd-ag-project-rd01pr-dmayn10")

library(sf)
library(DT)
library(arrow)
library(mapview)
library(fundiversity)
library(tidyverse)

rd_pathway <- "/Volumes/ritd-ag-project-rd01pr-dmayn10" # path to UCL research drive
```

# Introduction

## Narrative / Structure

-   **Ecological significance of forests**: Forests are biodiversity hotspots and provide vital ecosystem services, such as carbon storage, regulation of water cycles, and prevention of soil degradation.

-   **Successional dynamics:** Introduce forest succession, focus on the transition from pioneer to climax communities and the changes in species composition over time. Talks about stages of primary and secondary succession.

-   **Functional traits of trees:** Introduce the idea of functional traits and their role in determining tree species' strategies for growth, reproduction, and survival. Then make the link to why functional traits are more suited to understand and predict successional trajectories than taxonomic diversity metrics.

-   **Threats to (forest) biodiversity and why predictions are important**: Talk about the main threats to forests and why this may have wide ranging effects on global biodiversity and also human health (link to ecosystem services). Introduce observed and predicted effects of these threats (especially climate change) on forest dynamics, e.g., *forest phenology*.

## Objective and hypotheses

<mark> **Objective**: Here, we aim to get an insight into [*trends of change*] [*the effect of (covariate) on*] functional traits of trees over successional time. </mark>

<u> Potential research (sub-)question (don't mention in the text): </u>

-   **Main:** How do functional traits of trees change through forest successional time?

    -   **Sub 1:** How do [*covariates*] (biome, forest management, anthropogenic disturbances, climate change) affect [*traits* or *indices*]?
    -   **Sub 2:** How do [*covariates*] (biome, forest management, anthropogenic disturbances, climate change) affect proliferation strategies of trees (acquisitive vs conservative) at different stages of forest succession?
    -   **Sub 3:** Are successional trajectories deterministic or variable across different environments/biomes/gradients?\

<u> Potential hypotheses </u>:

-   Functional traits will shift from <mark> acquisitive traits </mark> in early succession to <mark> conservative traits </mark> in late succession (see [Lohbeck et al. 2013](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1890/12-1850.1) and [Faccion et al. 2021](https://ui.adsabs.harvard.edu/abs/2021FMDFE.27951815F/abstract)).

-   Successional trajectories in undisturbed environments (check forest management or anthropogenic disturbance) will be more deterministic, with predictable changes in functional traits, while those in disturbed environments will be more chaotic (see e.g., [Marcilio-Silver et al. 2016](https://link.springer.com/article/10.1556/168.2016.17.1.11)).

## Key concepts for the introduction

-   **Functional Traits and Ecological Strategies** - Discuss the trade-offs between fast growth (acqusitive traits) and resource conservation (conservative traits)

-   **Environmental Filtering** - Discuss the role of abiotic factors (e.g., water availability, light availability) in determining trait distributions.

-   **Community Assembly** - Discuss mechanisms driving changes in species composition and functional diversity over time (e.g., niche and dispersal assembly, maybe talk about *Hubble and Unified Neutral Theory* or the *Metacommunity theory*, especially in connection to deterministic vs chaotic trajectories). Generally, talk about distinguishing between predictable successional changes and those driven by random events.

-   **Resource-use strategies** - Trees may exhibit traits that aim towards resource acquisition (acquisitive traits: i.g., leafs with high area and short lifespan to increase photosynthetic rate, low wood density to increase growth. Willows (Salix spp.) and Poplars (Populus spp.) are high in these traits) or resource conservation (conservative traits: e.g., low leaf area size and life span to conserve more energy and maximise resource efficiency, high wood density to support structure and reduce damage)

# Data

```{r read trait data, echo = FALSE, warning = FALSE, message= FALSE}
# Read tree trait data 
traits <- read_csv("/Volumes/ritd-ag-project-rd01pr-dmayn10/trait_data/Estimated_trait_table_with_monos.csv")
```

There are two major data sources:

-   The **trait data** contains values for functional traits of global trees species (in total `r length(unique(traits$accepted_bin))` species). There are `observed` and `predicted` values, which correspond to empirically measured and imputed values. Predicted values are either imputed using environmental data, phylogenetic data, or both.\

```{r trait summaries, echo = FALSE, warning=FALSE, message=FALSE}
# Get trait summaries 
traits_summary <- traits %>%
  group_by(trait) %>%
  summarize(
    mean_pred_value = round(mean(pred_value, na.rm = TRUE), digits = 3),
    sd_pred_value   = round(sd(pred_value, na.rm = TRUE), digits = 3),
    min_pred_value  = round(min(pred_value, na.rm = TRUE), digits = 3),
    max_pred_value  = round(max(pred_value, na.rm = TRUE), digits = 3),
    mean_obs_value  = round(mean(obs_value, na.rm = TRUE), digits = 3), 
    sd_obs_value    = round(sd(obs_value, na.rm = TRUE), digits = 3),
    min_obs_value   = round(min(obs_value, na.rm = TRUE), digits = 3),
    max_obs_value   = round(max(obs_value, na.rm = TRUE), digits = 3))

# Plot table 
colnames(traits_summary) <- c("Trait", "Mean (pred)", "SD (pred)", "Min. (pred)", "Max. (pred)", "Mean (obs)", "SD (obs)", "Min. (obs)", "Max. (obs")
datatable(traits_summary, options = list(lengthMenu = c(5, 10, 20), pageLength = 5))
rm(traits_summary)
```

\
- The **FIA** data (Forest Inventory and Analysis) contains occurrences of tree species in the US arranged by states. For every state, there is an arrow file (import using `arrow::read_feather()`) containing the following columns:\

```{r FIA data description, echo = FALSE, warning=FALSE, message=FALSE}
# Get FIA summaries 
fia_summary <- tibble(
  Variable = c("pid", "PLT_CN", "CONDID", "SUBP", "TREE", "TREEID", "PREV_TREEID", 
"STATUSCD", "SPGRPCD", "SPCD", "DIA", "DIAHTCD", "CCLCD", "AGENTCD", 
"CPOSCD", "CLIGHTCD", "CARBON_AG", "STOCKING", "VOLCFGRS", "BHAGE", 
"TOTAGE", "MORTCD", "STANDING_DEAD_CD", "height", "TPA_UNADJ", 
"PID", "INVYR", "UNITCD", "STATECD", "COUNTYCD", "PLOT", "DESIGNCD", 
"KINDCD", "REMPER", "LAT", "LON", "MACRO_BREAKPOINT_DIA", "PLOT_STATUS_CD", 
"PREV_PLT_CN", "has_plot", "good_design", "forested", "CONDPROP_UNADJ", 
"PROP_BASIS", "COND_STATUS_CD", "OWNCD", "OWNGRPCD", "STDAGE", 
"FORINDCD", "FORTYPCD", "FORTYPCDCALC", "STDORGCD", "TRTCD1_P2A", 
"TRTCD1", "TRTCD2", "TRTCD3", "has_cond", "n", "mult_conds", 
"species", "accepted_bin", "tax_level", "managed", "alive", "Resolve_Biome", 
"biome", "canopy_pos", "ownership", "damage", "status", "biomass", 
"size_class", "PLOT_TYPE", "BPA", "USE", "STATE", "STATE_ABBRV", 
"max_year", "min_year", "LAST_YEAR", "FIRST_YEAR", "ll_id"),
  Description = c(
    "Plot ID", "Plot sequence number", "Condition class number", "Subplot number", 
    "Tree Number", "Tree ID", "Previous Tree ID", "Status code", "Species group code", 
    "Species code", "Diameter at breast height (inches)", 
    "Diameter height code indicating the location at which diameter was measured", 
    "Crown class code indicating the amount of sunlight received and the crown position within the canopy", 
    "Cause of death (agent) code", "Crown position code", "Crown light exposure code", 
    "Above ground carbon", "Stocking", "Gross cubic foot volume", "Breast height age", 
    "Total age", "Mortality code", "Standing dead code", "Tree height", 
    "Trees per acre (unadjusted)", "Plot identifier (duplicate)", "Inventory year", 
    "Unit code", "State code", "County code", "Plot number", "Design code", 
    "Sample kind code", "Re-measurement period", "Latitude", "Longitude", 
    "Macro plot breakpoint diameter", "Plot status code", "Previous plot CN", 
    "Plot presence indicator (?)", "Good design indicator (?)", "Forested indicator (?)", 
    "Condition proportion (unadjusted)", "Proportion basis", "Condition status code", 
    "Ownership code", "Ownership group code", "Stand age", "Forest industry code", 
    "Forest type code", "Calculated forest type code", "Stand origin code", 
    "Treatment code 1 P2A", "Treatment code 1", "Treatment code 2", "Treatment code 3", 
    "Condition presence indicator", "Count", "Multiple conditions indicator", "Species", 
    "Accepted binomial name", "Taxonomic level", "Managed indicator", "Alive indicator", 
    "Resolved biome", "Biome", "Canopy position", "Ownership", "Damage", "Status", 
    "Biomass", "Size class", "Plot type", "Basal area per acre", "Use code", "State", 
    "State abbreviation", "Maximum year", "Minimum year", "Last year observed", 
    "First year observed", "Longitude-latitude ID"))
datatable(fia_summary, options = list(lengthMenu = c(5, 20, 50, 100), pageLength = 5))
rm(fia_summary)
```

\

# Data preparation

## FIA data

We build our fia database by cleaning and summarizing the FIA data of every state to the species per plot level. This entails the following:

-   Filtering out dead trees, trees not identified to the species level, plots where the conditions is NOT 1 (see FIA description), non-fixed sampling designs, private ownership, and diameter values not measured at breast height.
-   Keeping only the most recent measurement per plot (?)
-   Scale numeric variables on tree level by trees per acre (TPA)
-   Summarize tree level variables to species per plot level using a weighted mean weighted by BA (basal area)

## Trait data

We complete our database by joining the FIA data per state with the trait data. This entails the following:

-   Selecting relevant traits for our analysis. These include traits that indicate resource use strategies and do not change drastically with tree age (<mark>confirm with Dan</mark>):

    -   **Wood density**: Measures the mass of wood per unit volume, indicating the strength and structural integrity of the wood.
    -   **Bark thickness**: The thickness of the tree's bark, providing protection against physical damage and pathogens.
    -   **Stem conduit diameter**: Is related to the efficiency of water transport within the tree. 
    -   **Seed dry mass**: The mass of a tree's seeds when dry, indicating investment in reproduction.
    -   **Leaf density**: The mass of a leaf per unit volume, indicative of structural robustness and resource investment.
    -   **Leaf N**: The concentration of nitrogen in leaves, reflecting the plant's investment in photosynthetic proteins.
    -   **Specific leaf area**: The leaf area per unit dry mass, indicative of the leaf's thickness and light-capturing efficiency.

    Some traits are retrieved from the FIA data:

    -   **mean_DIA**: The mean tree diameter per FIA plot.
    -   **mean_height**: The mean tree height per FIA plot.
    -   **sum_carbon_AG**: The sum of the carbon above the ground of trees in a FIA plot.
    -   **sum_BA**: The total basal area per FIA plot.
    -   **sum_biomass**: The total biomass area per FIA plot.
    
<br>

-   Filtering the trait data to only include traits imputed using phylogenetic data.

-   Filtering the trait data to only include species found in our respective FIA data

-   Classifying the trait values on a spectrum corresponding to resource use strategies:

    -   To quantify the (mean) resource-use strategy of each plot, we compute a composite score ranging from 0 to 1, where 0 indicates a highly acquisitive strategy and 1 indicates a highly conservative strategy. This is done using the means of the functional traits selected per plot:

        1.  **Normalization:** Each trait is normalized to a [0, 1] range using min-max normalization to ensure comparability. This is calculated as follows:

        $$ T_i (normalized) = \frac{T_{i} - T_{min}}{T_{max}- T_{min}} $$

        where T<sub>i</sub>​ is the value of the trait for plot <sub>i</sub>, T<sub>min</sub>​is the minimum value of the trait across all plots, and T<sub>max</sub>​ is the maximum value of the trait across all plots.

        2.  **Inversion for Acquisitive Traits**: Traits indicating a more acquisitive strategy (high values representing high acquisitiveness, e.g., leaf nitrogen and specific leaf area) were inverted using the formula:

        $$ T_i (inverted) = 1 - T_{normalized} $$

        3.  **Composite Score Calculation**: The resource-use strategy score for each plot was calculated as the mean of the normalized (and inverted) trait values. The composite resource-use strategy score *R*​ for plot *i* was calculated using the formula:

        $$ R = \frac{1}{n_{traits}} \sum_{i=1}^n T_{ij} $$

        where *n* is the number of traits included, T<sub>ij</sub> represents the normalized (and inverted) value of trait *j* for plot *i*.

        Based on the current selection of traits, we assign values as follows:

        -   **Wood Density** Higher wood density is generally associated with conservative strategies (higher resistance to damage, longer lifespan). Assign higher values (close to 1) for high wood density.
        -   **Bark Thickness:** Thicker bark is associated with more protection and conservation of resources. Assign higher values (close to 1) for thicker bark.
        -   **Conduit diam.:** Larger conduit diameters generally allow for higher rates of water transport, which can support faster growth and higher photosynthetic rates and therefore acquisitive.
        -   **Seed Dry Mass:** Larger seed mass is often related to higher investment per offspring, indicating a conservative strategy. Assign higher values (close to 1) for higher seed mass.
        -   **Leaf Density:** Denser leaves may imply more investment in structural tissues, hence conservative. Assign higher values (close to 1) for higher leaf density.
        -   **Leaf N (Nitrogen Content):** Higher nitrogen content typically indicates higher photosynthetic rates and faster growth, associated with acquisitive strategies. Assign lower values (close to 0) for high nitrogen content.
        -   **Specific Leaf Area:** Higher SLA indicates thinner leaves with less investment in structural tissues, associated with acquisitive strategies. Assign lower values (close to 0) for high SLA.

        See the following literature:

        -   [Poorter et al. 2008](https://esajournals.onlinelibrary.wiley.com/doi/10.1890/07-0207.1) found that wood density and seed volume are significant predictors of growth and mortality, with denser wood and larger seeds being associated with slower growth and longer survival, typical of conservative traits.
        -   [Visser et al. 2016](https://besjournals.onlinelibrary.wiley.com/doi/10.1111/1365-2435.12621) demonstrated that wood density, seed mass, and adult stature significantly predict various demographic rates, supporting their classification as conservative trait
        -   [Plaine et al. 2015](https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/1365-2745.12401) indicated that traits like specific leaf area and wood density are linked to growth rates, with high SLA linked to fast growth (acquisitive) and high wood density linked to slower growth (conservative)

## Functional diversity indices

Based on the species per plots and the selected functional traits, we calculate:

-   **Functional divergence**: Measures how species diverge in their functional traits from the mean trait value of the community, indicating niche differentiation and how species use the available niche space differently.
-   **Functional dispersion**: Quantifies the spread of species in a community based on their functional traits, reflecting the overall trait variation and dissimilarity within the community.
-   **Functional evenness**: Assesses how evenly functional traits are distributed among species in a community, indicating the uniformity of trait abundances and resource use.

Note: These are calculated using the species means per plot. Trait values are scaled before index calculations with a mean of zero and a standard deviation of one (z-score).

## Database

Finally, we summarise the entire dataset to the plot (PID) level, where eventually one row is one plot. We summarise the mean, median, standard deviation, and interquartile range. For every plot we then know the biome, inventory year, ownership, if it is managed, the location ID, the summary statistics of the functional traits wood density, lead nitrogen, bark thickness, seed dry mass, leaf density, specific leaf area, diameter, and basal area per acre. We also know the sum of the carbon above the ground and the stand age. For each plot community we also know the functional dispersion, diversion, and evenness.

The entire preparation pipelines is defined in the following function:

```{r preparation pipeline, echo = TRUE}

prep_fia <- function(arrow_path, trait_data){
	
# Make sure libraries are open
require(arrow)
require(tidyverse)

########## Run preparation pipeline on state FIA data ##########

# Read state FIA file using pathname 
fia <- arrow::read_feather(arrow_path) %>%
	
	## Filtering (while making sure formats are correct) 
	
	# Dead: No dead trees 
	mutate(alive = as.integer(alive)) %>%
	filter(alive == 1) %>%
	
	# Taxonomy: Only species level trees
	mutate(tax_level = as.integer(tax_level)) %>%
	filter(tax_level == 2) %>%
	
	# Condition: Only plots with condition 1
	mutate(CONDID = as.integer(CONDID)) %>%
	filter(CONDID == 1) %>%
	mutate(mult_conds = as.integer(mult_conds)) %>%
	filter(mult_conds == 0) %>%
	
	# Design: Only fixed design (plus with macros to add CA, OR, WA)
	mutate(DESIGNCD = as.integer(DESIGNCD)) %>%
	filter(DESIGNCD %in% c(505, 501, 1)) %>%
	
	# Ownership: Disregard private ownerships
	mutate(OWNGRPCD = as.integer(OWNGRPCD)) %>%
	filter(OWNGRPCD != 40 ) %>%
	
	# DIA: Only diameter values measured at breast height
	mutate(DIAHTCD = as.integer(DIAHTCD)) %>%
	filter(DIAHTCD == 1) %>%
	
	## Database management 
	
	# Keep only the most recent observation per PID (for now)
	group_by(PID) %>%
	filter(INVYR == max(INVYR)) %>%
	ungroup() %>%
	
	# Remove columns with no variation and set format 
	select(accepted_bin, TREEID, PID, SUBP, PLOT_TYPE, TPA_UNADJ, INVYR, STDAGE, biome,
				 DIA, CARBON_AG, STOCKING, BPA, biomass, height, size_class, canopy_pos, ownership,
				 managed, TRTCD1, TRTCD2, USE, ll_id, min_year, max_year) %>%
	
	# Calculate Basal Area using the DIA 
	mutate(BA = pi * (DIA/2)^2) %>%
	
	## Scale numeric tree level variables by TPA
	
	mutate(across(
		c(CARBON_AG, BA, biomass), ~ . * TPA_UNADJ, .names = "{col}_scaled")) %>%
	select(-TPA_UNADJ, -CARBON_AG, -BA, -biomass)

## Summarize to species per plot level

# First do a safety check to make sure there are no more than 1 levels to PID per 
# grouping variable (excluding accepted_bin): 

grouping_vars <- c("biome", "INVYR", "ownership", "managed", "TRTCD1", "TRTCD2", "ll_id")

for (var in grouping_vars) {
	level_check <- fia %>%
		group_by(PID) %>%
		summarise(levels_count = n_distinct(.data[[var]])) %>%
		filter(levels_count > 1)
	
	if (nrow(level_check) > 0) {
		stop(paste("Error: More than one level per PID in grouping variable", var, "detected!"))
	}
}

# Now summarize data to species per plot level

fia <- fia %>%
	group_by(PID, accepted_bin, biome, INVYR, ownership, managed, ll_id) %>%
	summarize(
		# Means for DIA and height 
		mean_DIA = mean(DIA, na.rm = TRUE),
		mean_height = mean(height, na.rm = TRUE),
		# Sums for the rest
		sum_BPA = sum(BPA, na.rm = TRUE),
		sum_BA = sum(BA_scaled, na.rm = TRUE),
		sum_biomass = sum(biomass_scaled, na.rm = TRUE),
		sum_CARBON_AG = sum(CARBON_AG_scaled, na.rm = TRUE),
		mean_STDAGE = mean(STDAGE, na.rm = TRUE),
		.groups = 'drop') %>%
	
	## Finish and set format
	
	ungroup() %>%
	as_tibble()

########## Join FIA data with trait data ##########

dat <- trait_data %>%
	
	# Keep only relevant traits to the analysis and 
	# those that do not change drastically with tree age 
	filter(trait_short %in% c("Wood density", "Bark thickness", "Seed dry mass", 
														"Leaf density", "Leaf P", "Specific leaf area", "Conduit diam.")) %>%
	
	# Filter by imputation: only phylogenetically imputed values 
	filter(fit == "phy") %>% 
	
	# Filter by species present in FIA data 
	filter(accepted_bin %in% fia$accepted_bin) %>%
	
	# Clean trait column (replace white spaces)
	mutate(trait_short = gsub(" ", "_", trait_short),
				 trait_short = tolower(trait_short)) %>%
	
	# Keep only species and (predicted) trait
	select(accepted_bin, trait_short, pred_value) %>%
	
	# Pivot wider
	pivot_wider(names_from = trait_short, 
							values_from = pred_value,
							names_sep = "_") %>%
	
	# Un-group and set format 
	ungroup() %>%
	as_tibble() %>%
	
	# Join traits with FIA data 
	right_join(fia, by = "accepted_bin")

########## Calculate functional diversity indices and build database ##########

# Summarise and scale trait data and format as matrix to calculate indices
fun_traits <- dat %>%
	
	# Keep relevant columns 
	select(accepted_bin, wood_density, leaf_p, bark_thickness, seed_dry_mass, conduit_diam.,
				 leaf_density, specific_leaf_area, mean_DIA, sum_BPA, mean_height, sum_CARBON_AG, sum_biomass, sum_BA) %>%
	
	# Group by species
	group_by(accepted_bin) %>%
	
	# Summarise traits to plot level 
	summarise(
		
		# Means weighted by BA
		dia = mean(mean_DIA),
		height = mean(mean_height),
		
		# Means weighted by BA
		wood_density = weighted.mean(wood_density, sum_BA),
		bark_thickness = weighted.mean(bark_thickness, sum_BA),
		conduit_diam = weighted.mean(conduit_diam., sum_BA),
		leaf_p = weighted.mean(leaf_p, sum_BA),
		specific_leaf_area = weighted.mean(specific_leaf_area, sum_BA),
		leaf_density = weighted.mean(leaf_density, sum_BA),
		seed_dry_mass = weighted.mean(seed_dry_mass, sum_BA),
		
		# Sums 
		carbon_ag = sum(sum_CARBON_AG),
		biomass = sum(sum_biomass),
		BA = sum(sum_BA),
		
		.groups = 'drop') %>%
	
	# Drop missing values 
	filter(complete.cases(.)) %>%
	
	# Format as matrix 
	column_to_rownames(var = "accepted_bin") %>%
	as.matrix() %>%
	
	# Scale values
	scale()

# Summarise plot data and format as presence/absence matrix 
fun_plots <- dat %>%
	# Keep only relevant columns 
	select(accepted_bin, PID) %>%
	# Set presence to 1
	mutate(present = 1) %>%
	# Keep only species found in trait data 
	filter(accepted_bin %in% rownames(fun_traits)) %>%
	# Pivot to wider 
	pivot_wider(names_from = accepted_bin, values_from = present, values_fill = list(present = 0)) %>%
	# Format as matrix 
	column_to_rownames(var = "PID") %>%
	as.matrix()

# Functions to normalize trait vectors to a range [0, 1] used for resource use score
normalize <- function(x) {
	(x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}

# The invert for acquisitive traits 
invert_normalized <- function(x) {
	return(1 - x)
}

## Summarise to plot level, calculate resource use score, and functional indices

# Summarise data to plot level 
fia <- dat %>%
	# Define grouping variables that are constant for every plot 
	group_by(PID, biome, INVYR, ownership, managed, ll_id) %>%
	
	# Summarise to plot levels by means (using weighted means by BA), SD, and sums
	summarise(
		
		# Stand age (this should be constant for every plot)
		standage = mean(mean_STDAGE),
		
		# Wood density 
		wmean_wood_density = weighted.mean(wood_density, sum_BA, na.rm = TRUE),
		sd_wood_density = sd(wood_density, na.rm = TRUE),
		
		# Bark thickness
		wmean_bark_thickness = weighted.mean(bark_thickness, sum_BA, na.rm = TRUE),
		sd_bark_thickness = sd(bark_thickness, na.rm = TRUE),
		
		# Conduit diameter
		wmean_conduit_diam = weighted.mean(conduit_diam., sum_BA, na.rm = TRUE),
    sd_conduit_diam = sd(conduit_diam., na.rm = TRUE),

		# Leaf pottassium 
		wmean_leaf_p = weighted.mean(leaf_p, sum_BA, na.rm = TRUE),
		sd_leaf_p = sd(leaf_p, na.rm = TRUE),

		# Leaf density
		wmean_leaf_density = weighted.mean(leaf_density, sum_BA, na.rm = TRUE),
		sd_leaf_density = sd(leaf_density, na.rm = TRUE),

		# Specific leaf area
		wmean_specific_leaf_area = weighted.mean(specific_leaf_area, sum_BA, na.rm = TRUE),
		sd_specific_leaf_area = sd(specific_leaf_area, na.rm = TRUE),

		# Seed dry mass
		wmean_seed_dry_mass = weighted.mean(seed_dry_mass, sum_BA, na.rm = TRUE),
		sd_seed_dry_mass = sd(seed_dry_mass, na.rm = TRUE),

		# Diameter
		mean_dia = mean(mean_DIA, na.rm = TRUE),
		sd_dia = sd(mean_DIA, na.rm = TRUE),
		
		# Height 
		wmean_height = weighted.mean(mean_height, sum_BA, na.rm = TRUE),
		sd_height = sd(mean_height, na.rm = TRUE),

		# Above the ground carbon (SUM)
		sum_carbon_ag = sum(sum_CARBON_AG),

    # Biomass (SUM)
    sum_biomass = sum(sum_biomass),

    # Basal area (SUM)
    basal_area = sum(sum_BA),
		.groups = 'drop') %>%
	
	ungroup() %>%
	
	## Approximate resource use strategy per plot
	
	mutate(
		# Calculate the composite resource-use strategy score
		resource_use_score = (
			  # Wood density indicating conservative 
		  	normalize(wmean_wood_density) + 
				# Bark thickness indicating conservative
				normalize(wmean_bark_thickness) +
				# Conduit diameter indicating acquisitive
				invert_normalized(normalize(wmean_conduit_diam)) +
				# Leaf potassium indicating acquisitive
				invert_normalized(normalize(wmean_leaf_p)) +
				# Leaf density indicating conservative
				normalize(wmean_leaf_density) +
				# Specific leaf area indicating acquisitive
				invert_normalized(normalize(wmean_specific_leaf_area)) +
				# Seed dry mass indicating conservative
				normalize(wmean_seed_dry_mass)) / 7 ) %>%
	
	## Calculate functional diversity indices per plot
	
	# Functional divergence 
	left_join(fd_fdiv(fun_traits, fun_plots) %>% rename(PID = site), by = "PID") %>%
	rename(fun_div = FDiv) %>%
	
	# Functional dispersion
	left_join(fd_fdis(fun_traits, fun_plots) %>% rename(PID = site), by = "PID") %>%
	rename(fun_disp = FDis) %>%
	
	# Functional evenness
	left_join(fd_feve(fun_traits, fun_plots) %>% rename(PID = site), by = "PID") %>%
	rename(fun_even = FEve) %>%
	
	# Set format
	mutate_all(~ ifelse(is.nan(.), NA, .)) %>%
	as_tibble()

  	
	
 ## Return cleaned FIA data 
 return(fia)
}

```

This is then passed to a vector containing the pathways to all FIA files (arrow per state).

*NOTE: CA, OR, and WA do not yield data because there are no `good_design == 1` plots. For now, we keep all filters intact and skip FIA files where no data was retrieved.*

```{r prep FIA data, echo=TRUE, warning = FALSE, message=FALSE}

# Wrap the FIA prep pipeline with safely
prep_fia_safe <- safely(prep_fia, otherwise = NA)

# Access FIA pathways 
fia_data <- tibble(pathway = list.files(path = "/Volumes/ritd-ag-project-rd01pr-dmayn10/fia_data/cleaned_state_data",
									 full.names = TRUE)) %>%
	# Extract state name from file name 
	mutate(state = str_extract(pathway, "(?<=/FIA_)[^\\.]+"))

# Pass data through preparation pipeline while skipping problematic files 
fia_data <- fia_data %>%
  mutate(result = map(pathway, ~ prep_fia_safe(.x, traits)),
         data = map(result, "result"),
         has_problems = map_lgl(result, ~ !is.null(.x$error))) %>%
  select(state, data, has_problems)

# Get available data and write file 
fia_clean <- fia_data %>%
  unnest(data) %>%
	write_csv(file = paste0("/Volumes/ritd-ag-project-rd01pr-dmayn10/merlin/data/fia_traits/plotlvl_traits_", Sys.Date(),".csv"))
```

As a preliminary exploration, we can examine the relationship between our traits and the stand age of the plots as a simple proxy for successional stage (cut-off at 300 years):

```{r remove weird values, echo=FALSE}
fia_clean <- fia_clean %>%
	filter(standage < 300)
```

## Functional Diversity Indices and resource use score

```{r plot indices, echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10, fig.cap = "Functional diversity indiced vs stand age"}

fia_clean %>%
	# Keep PID, stand age, and mean traits
	select(PID, standage, fun_div, fun_disp, fun_even, resource_use_score) %>%
	# Pivot longer
	pivot_longer(
	cols = c(fun_div, fun_disp, fun_even, resource_use_score),       
	names_to = "Index",               
	values_to = "value") %>%
	filter(value > 0 & value < 1) %>%
	# Plot indices against stand age 
	ggplot(aes(x=standage, y=value, group = Index)) +
	geom_hex() +
  scale_fill_viridis_c() +
	geom_smooth(method = "lm", colour = "red") +
	geom_smooth(method = "lm", colour = "darkred") +
	facet_wrap(~Index, scales = "free") +
	theme_bw() +
	theme(legend.position = "none") +
	ggtitle("Functional diversity indices and resource use score") +
	xlab("Stand age (years)") +
	ylab("Index values")

```

This becomes more interesting when look at per biome: 

```{r user per biome, echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10, fig.cap = "Resource use vs stand age per biome"}
fia_clean %>%
	# Keep PID, stand age, and mean traits
	select(PID, standage,resource_use_score, biome) %>%
	filter(complete.cases(.)) %>%
	# Plot indices against stand age 
	ggplot(aes(x=standage, y=resource_use_score, group = biome)) +
	geom_hex() +
	scale_fill_viridis_c() +
	geom_smooth(method = "lm", colour = "red") +
	geom_smooth(method = "lm", colour = "darkred") +
	facet_wrap(~biome, scales = "free") +
	theme_bw() +
	theme(legend.position = "none") +
	ggtitle("Resource use score vs stand age per biome") +
	xlab("Stand age (years)") +
	ylab("Index values")
```

basal_area, sum_biomass, sum_carbon_ag

## Total Trait Sums

```{r plot means, echo = FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=4, fig.cap = "Trait means vs stand age"}
fia_clean %>%
	# Keep PID, stand age, and mean traits
	select(PID, standage, basal_area, sum_biomass, sum_carbon_ag) %>%
	# Pivot longer
	pivot_longer(
	cols = c(basal_area, sum_biomass, sum_carbon_ag),       
	names_to = "Trait",               
	values_to = "value") %>%
	# Plot mean traits against stand age 
	ggplot(aes(x=standage, y=value, group = Trait)) +
	geom_hex() +
  scale_fill_viridis_c() +
	geom_smooth(method = "lm", colour = "red") +
	facet_wrap(~Trait, scales = "free") +
	theme_bw() +
	theme(legend.position = "none") +
	ggtitle("Trait sums") +
	xlab("Stand age (years)") +
	ylab("Values")
```


## Weighted Trait Means

```{r plot means, echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10, fig.cap = "Trait means vs stand age"}
fia_clean %>%
	# Keep PID, stand age, and mean traits
	select(PID, standage, wmean_wood_density, wmean_bark_thickness, wmean_leaf_p, wmean_conduit_diam,
				 wmean_leaf_density, wmean_specific_leaf_area, wmean_seed_dry_mass, wmean_height, mean_dia) %>%
	# Pivot longer
	pivot_longer(
	cols = c(wmean_wood_density, wmean_bark_thickness, wmean_conduit_diam, wmean_leaf_p, 
					 wmean_leaf_density, wmean_specific_leaf_area, wmean_seed_dry_mass, wmean_height, mean_dia),       
	names_to = "Trait",               
	values_to = "value") %>%
	# Plot mean traits against stand age 
	ggplot(aes(x=standage, y=value, group = Trait)) +
	geom_hex() +
  scale_fill_viridis_c() +
	geom_smooth(method = "lm", colour = "red") +
	facet_wrap(~Trait, scales = "free") +
	theme_bw() +
	theme(legend.position = "none") +
	ggtitle("Trait means") +
	xlab("Stand age (years)") +
	ylab("Values")
```

## Trait standard deviations

```{r plot sd, echo = FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=10, fig.cap = "Trait sd vs stand age"}
fia_clean %>%
	# Keep PID, stand age, and sd traits
	select(PID, standage, sd_wood_density, sd_bark_thickness, sd_conduit_diam, sd_leaf_p, 
				 sd_leaf_density, sd_specific_leaf_area, sd_seed_dry_mass, sd_dia, sd_height,  sum_carbon_ag) %>%
	# Pivot longer
	pivot_longer(
	cols = c(sd_wood_density, sd_bark_thickness, sd_leaf_p, sd_conduit_diam,
					 sd_leaf_density, sd_specific_leaf_area, sd_seed_dry_mass, sd_dia, sd_height),       
	names_to = "Trait",               
	values_to = "value") %>%
	# Plot sd traits against stand age 
	ggplot(aes(x=standage, y=value, group = Trait)) +
	geom_hex() +
  scale_fill_viridis_c() +
	geom_smooth(method = "lm", colour = "red") +
	geom_smooth(method = "lm", colour = "forestgreen") +
	facet_wrap(~Trait, scales = "free") +
	theme_bw() +
	theme(legend.position = "none") +
	ggtitle("Trait sd") +
	xlab("Stand age (years)") +
	ylab("Values")
```

## Spatial distribution

We can also examine the spatial distribution (and density) of plots using a heatmap (excluding Hawaii and Alaska):

```{r map fia plots, echo=FALSE, warning=FALSE, message=FALSE}

library(rnaturalearth) 
library(rnaturalearthdata) 

read_csv("/Volumes/ritd-ag-project-rd01pr-dmayn10/fia_data/lat_long_lookup.csv") %>%
	right_join(
		select(fia_clean, PID, ll_id, state)
		) %>%
	filter(LON <= 0 & LON > -125 & LAT > 20) %>%
	
	ggplot(aes(x = LON, y = LAT)) +
	coord_sf(crs = 4326) +
  geom_bin2d(bins = 500) + 
  scale_fill_gradient(low = "blue", high = "red") +
  theme_classic() +
	theme(legend.position = c(.075, .22)) +
  labs(title = "Heat Map of FIA plots",
       x = "Longitude",
       y = "Latitude",
       fill = "Count")
```
