---
title: "<small><i>MSc Ecology and Data Science Thesis</i></small> <br> Functional traits of trees through forest successional time <br> - </br> <small> Traits Analysis - Random Forest </br>"
author: "Merlin Weiss"
date: "Status: `r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github
    output_dir: "/Users/serpent/Documents/MSc/Thesis/Markdown"
---

```{r setup, include=FALSE}

rm(list = ls()) # clean environment 

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Volumes/ritd-ag-project-rd01pr-dmayn10/merlin")

library(DT)
library(caret)
library(ranger)
library(pdp)
library(fastshap)
library(doParallel)
library(ggbeeswarm)
library(tidyverse)

rd_pathway <- "/Volumes/ritd-ag-project-rd01pr-dmayn10" # path to UCL research drive
data <- read_csv("/Volumes/ritd-ag-project-rd01pr-dmayn10/merlin/data/fia_traits/traits_rf_clean.csv")
data <- data %>% sample_n(500) # only for code development 
```

# Objectives

1. Understand the impact of stand age on forest traits.
2. Examine how this relationship varies across different ecoregions and climate conditions.
3. Interpret model results using Shapley values to understand feature importance.

# Outline of the Approach

1. Data Preparation:
    - Ensure the dataset is ready with all necessary variables: traits, stand age, ecoregion, climate variables, and other covariates.
    - Split the data into training and testing sets.
2. Hyperparameter Tuning:
    - Perform grid search combined with cross-validation to find the optimal hyperparameters (mtry, num.trees, min.node.size).
3. Model Fitting with Optimal Hyperparameters:
    - Use the optimal hyperparameters to fit random forest models for each trait.
    - Ensure stand age is included as a mandatory predictor in all models.
4. Interpretation with Shapley Values:
    - Calculate Shapley values for feature importance using the fastshap package.
    - Visualize the importance of stand age and other covariates across different models.
5. Analysis Across Ecoregions and Climate Conditions:
    - Stratify data by ecoregion and temperature quantiles.
    - Fit separate models for each stratum using the optimal hyperparameters and compare results.
    
## Code implementation

Split data into train and test data: 
```{r prep and split data, echo=TRUE}
# Split into training and test sets
set.seed(42)
train_indices <- sample(1:nrow(data), size = 0.8 * nrow(data))
train_data <- data[train_indices, ]
test_data  <- data[-train_indices, ]
```

Define traits and covariates: 
```{r defina variables }
# Define the traits and covariates
traits <- c("wmean_wood_density", "wmean_bark_thickness", "wmean_conduit_diam", "wmean_leaf_n", 
						"wmean_specific_leaf_area", "wmean_seed_dry_mass", "wmean_shade_tolerance", "wmean_height")

covariates <- c("standage", "annual_mean_temperature", "annual_precipitation", 
								"temperature_seasonality", "mean_diurnal_range", 
								"min_temperature_of_coldest_month", "max_temperature_of_warmest_month", 
								"elevation", "pop_density", "sand_content_015cm", 
								"soil_ph_015cm", "water_capacity_015cm")
```


Find best models using a grid search on hyperparameters and cross validation 

```{r model training}
# Define a grid of hyperparameters including mtry and min.niode.size
hyper_grid <- expand.grid(
	mtry = seq(2, length(covariates), by = 2),
	min.node.size = c(1, 5, 10),
	splitrule = "variance")

# Perform cross-validation, find the best hyperparameters, capture performance metrics
tune_rf_model <- function(trait, data, covariates, hyper_grid) {
	
	# Define model formula 
	formula <- as.formula(paste(trait, "~", paste(covariates, collapse = " + ")))
	
	# Set training framework 
	train_control <- trainControl(method = "cv", number = 5, savePredictions = "final", summaryFunction = defaultSummary)
	
	# Train models 
	model <- train(
		formula, data = data,
		method = "ranger",
		trControl = train_control,
		tuneGrid = hyper_grid,
		importance = "permutation"
	)
	
	# Get performance metrics
	results <- model$results
	best_tune <- model$bestTune
	results <- results %>%
		mutate(trait = trait, 
					 best = rowSums(sapply(names(best_tune), 
					 											function(param) results[[param]] == best_tune[[param]])) == length(best_tune))
	
	return(list(model = model, results = results))
}

# Perform hyperparameter tuning for each trait and get performance metrics
tuned_models <- lapply(traits, function(trait) tune_rf_model(trait, train_data, covariates, hyper_grid))

# Extract the best models and performance metrics
best_models <- lapply(tuned_models, function(x) x$model$finalModel)
performance_metrics <- do.call(rbind, lapply(tuned_models, function(x) x$results))
```

Examine performance metrics: 
```{r show metrics, echo=FALSE}
datatable(performance_metrics, options = list(lengthMenu = c(5, 10, 20, 50, 100, 150), pageLength = 5))
```

In this table: 

- **RMSE: Root Mean Squared Error**. This is a measure of the average deviation of the predicted values from the actual values. It is calculated as the square root of the average of squared differences between predicted and actual values. Lower values indicate better model performance.
- **R-squared**. This is a measure of how well the model’s predictions approximate the actual data. It represents the proportion of variance in the dependent variable that is predictable from the independent variables. Values range from 0 to 1, with higher values indicating better model performance.
- **MAE: Mean Absolute Error**. This is the average of the absolute differences between predicted and actual values. It provides a measure of how far the predictions are from the actual values on average. Lower values indicate better model performance.
- **RMSESD: Standard Deviation of Root Mean Squared Error**. This measures the variability of the RMSE across different cross-validation folds. Lower values indicate more consistent model performance.
- **R-squaredSD: Standard Deviation of R-squared**. This measures the variability of the R-squared values across different cross-validation folds. Lower values indicate more consistent model performance.
- **MAESD: Standard Deviation of Mean Absolute Error**. This measures the variability of the MAE across different cross-validation folds. Lower values indicate more consistent model performance.

Next: Make predictions and calculate SHAPLEY values 

```{r shapley analysis, echo=TRUE, fig.width=15, fig.height=8}

# Define prediction function for fastshap
predict_fn <- function(object, newdata) {
	predict(object, data = newdata)$predictions
}

# Calculate Shapley values
registerDoParallel(detectCores() - 1)
shap_values <- lapply(best_models, function(model) {
	fastshap::explain(model, X = test_data[, covariates, drop = FALSE], 
										pred_wrapper = predict_fn, nsim = 100, parallel = TRUE)
})

# Define a function to plot Shapley values
plot_shap_values <- function(shap_values, covariates, trait) {
	# Reshape Shapley values to a long format
	shap_df <- as.data.frame(shap_values)
	shap_df$id <- 1:nrow(shap_df)
	shap_long <- tidyr::gather(shap_df, key = "feature", value = "shap", -id)
	
	# Ensure the features match the covariates
	shap_long$feature <- rep(covariates, each = nrow(shap_df))
	
	# Plotting
	ggplot(shap_long, aes(x = feature, y = shap, color = shap)) +
		geom_quasirandom(alpha = 0.5) +
		scale_color_viridis_c(option = "viridis") +
		coord_flip() +
		labs(title = paste(trait), x = "Feature", y = "Shapley Value") +
		theme_bw() +
		theme(text = element_text(family = "Palatino"),
					legend.position = "none")
}

plots <- lapply(seq_along(traits), function(i) plot_shap_values(shap_values[[i]], covariates, traits[i]))
combined_plot <- cowplot::plot_grid(plotlist = plots, ncol = 4, nrow = 2)
print(combined_plot)
```

